---
title: "Housing affordability: Canterbury vs NZ"
output:
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(readxl)
library(zoo)
library(sandwich)
library(lmtest)
# car is called as car::vif(): loading it would mask dplyr::recode, used in the plots

covid         <- as.Date("2020-03-01")  # first post-COVID month; the single break date used throughout
common_window <- TRUE                   # TRUE: NZ and Canterbury full-sample models use identical months
```

## Load and prepare data

```{r load}
raw <- read_excel("Time_series_data.xlsx", sheet = "Combined")
names(raw) <- str_squish(names(raw))   # strips trailing/double spaces

df <- raw %>%
  transmute(
    date          = as.Date(Date),
    nz_price      = `NZ Median sale price`,
    cant_price    = `Canterbury Median sale price`,
    nz_rent       = `Median Rent`,
    cant_rent     = `Canterbury Median Rent`,
    nz_low_rent   = `Lower Quartile Rent`,
    cant_low_rent = `Canterbury Lower Quartile Rent`,
    rate_1y       = `1 Year Mortgage Rate`,
    rate_2y       = `2 Year Mortgage Rate`,
    ocr           = OCR,
    income_nz     = `Mean Income`,              # per person
    income_cant   = `Mean Income Canterbury`,   # per person
    unemp         = `Unemployment Rate`,
    cpi           = CPI,
    pop_nz        = Population,
    pop_cant      = `Canterbury Population`,
    lvr_total     = `LVR Total New Commitments (NZDm)`,
    lvr_above80   = `LVR Above 80% Commitments (NZDm)`,
    consents_nz   = `Dwelling Unit Consents`,
    consents_cant = `Canterbury Dwelling Unit Consents`,
    vol_nz        = `Volume Sold`,
    vol_cant      = `Volume Sold Canterbury`,
    net_mig_nz    = `Net migration`,
    bright_line   = `Bright Line Test Period`,   # holding period in force (years)
    invent_nz     = `Weeks of Inventory`,
    invent_cant   = `Weeks of Inventory Canterbury`
  )

stopifnot(all(as.numeric(diff(df$date)) %in% 28:31))   # monthly grid, no gaps
```

```{r derive}
# Linear interpolation between observed points; na.rm = FALSE means no extrapolation
interp_vars <- c("income_nz", "income_cant", "unemp", "cpi", "pop_nz", "pop_cant")
df <- df %>% mutate(across(all_of(interp_vars), ~ na.approx(.x, na.rm = FALSE)))

# Monthly P&I payment on an 80% loan, as % of annual income
pmt_share <- function(price, rate_pct, income, ltv = 0.8, years = 30) {
  r <- rate_pct / 100 / 12
  n <- years * 12
  price * ltv * r / (1 - (1 + r)^(-n)) * 12 / income * 100
}

df <- df %>%
  mutate(
    # affordability outcomes (2-year rate main, 1-year sensitivity)
    pti_nz       = nz_price / income_nz,
    pti_cant     = cant_price / income_cant,
    serv_nz_2y   = pmt_share(nz_price,   rate_2y, income_nz),
    serv_cant_2y = pmt_share(cant_price, rate_2y, income_cant),
    serv_nz_1y   = pmt_share(nz_price,   rate_1y, income_nz),
    serv_cant_1y = pmt_share(cant_price, rate_1y, income_cant),
    rti_nz       = nz_rent   * 52 / income_nz   * 100,
    rti_cant     = cant_rent * 52 / income_cant * 100,

    # regressors
    lvr_share       = lvr_above80 / lvr_total * 100,
    net_mig_nz_k    = net_mig_nz / 1000,
    pop_growth_nz   = 100 * (pop_nz   / dplyr::lag(pop_nz, 12)   - 1),
    pop_growth_cant = 100 * (pop_cant / dplyr::lag(pop_cant, 12) - 1),

    consents_12m_nz       = rollsumr(consents_nz,   12, fill = NA),
    consents_12m_cant     = rollsumr(consents_cant, 12, fill = NA),
    consents_per1000_nz   = consents_12m_nz   / pop_nz   * 1000,
    consents_per1000_cant = consents_12m_cant / pop_cant * 1000,

    vol_12m_nz   = rollsumr(vol_nz,   12, fill = NA),
    vol_12m_cant = rollsumr(vol_cant, 12, fill = NA),
    vol_yoy_nz   = 100 * (vol_12m_nz   / dplyr::lag(vol_12m_nz, 12)   - 1),
    vol_yoy_cant = 100 * (vol_12m_cant / dplyr::lag(vol_12m_cant, 12) - 1),
    
    invent_nz = invent_nz,
    invent_cant = invent_cant
  )
```

## Sanity checks

```{r checks}
# Coverage of every variable
df %>%
  pivot_longer(-date) %>%
  filter(!is.na(value)) %>%
  group_by(name) %>%
  summarise(n = n(), first = min(date), last = max(date), .groups = "drop") %>%
  print(n = Inf)

# Constructed measures vs the official annual points
# (levels differ: per-person vs household income; compare trends)
official <- raw %>%
  transmute(date = as.Date(Date),
            off_pti_nz   = `New Zealand House value to income`,
            off_pti_cant = `Canterbury Region House value to income`,
            off_ser_nz   = `New Zealand Mortgage payment proportion of income`,
            off_ser_cant = `Canterbury Region Mortgage payment proportion of income`) %>%
  filter(!is.na(off_pti_nz)) %>%
  inner_join(df %>% select(date, pti_nz, pti_cant, serv_nz_2y, serv_cant_2y,
                           serv_nz_1y, serv_cant_1y), by = "date") %>%
  drop_na()

tibble(
  measure  = c("PTI", "Serv 2y", "Serv 1y"),
  cor_nz   = c(cor(official$off_pti_nz, official$pti_nz),
               cor(official$off_ser_nz, official$serv_nz_2y),
               cor(official$off_ser_nz, official$serv_nz_1y)),
  cor_cant = c(cor(official$off_pti_cant, official$pti_cant),
               cor(official$off_ser_cant, official$serv_cant_2y),
               cor(official$off_ser_cant, official$serv_cant_1y))
) %>% mutate(across(where(is.numeric), ~ round(.x, 2)))
```

## Descriptives

```{r outcomes, fig.height = 8}
outcomes <- df %>%
  select(date, pti_nz, pti_cant, serv_nz = serv_nz_2y, serv_cant = serv_cant_2y,
         rti_nz, rti_cant) %>%
  pivot_longer(-date, names_to = c("measure", "region"), names_pattern = "(.*)_(nz|cant)") %>%
  filter(!is.na(value)) %>%
  mutate(region  = if_else(region == "nz", "NZ", "Canterbury"),
         measure = factor(measure, levels = c("pti", "serv", "rti"),
                          labels = c("Price-to-income (x)",
                                     "Mortgage payment (% of income, 2-yr rate)",
                                     "Rent-to-income (%)")))

ggplot(outcomes, aes(date, value, colour = region)) +
  geom_line() +
  geom_vline(xintercept = covid, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ measure, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = NULL, colour = NULL,
       title = "Affordability: Canterbury vs NZ (per-person income)",
       caption = paste0("Dashed line: ", format(covid, "%B %Y"),
                        ". Income interpolated linearly from annual data.")) +
  theme_minimal() + theme(legend.position = "top")
```

```{r drivers, fig.height = 8}
by_region <- df %>%
  select(date, consents_per1000_nz, consents_per1000_cant,
         pop_growth_nz, pop_growth_cant, vol_yoy_nz, vol_yoy_cant) %>%
  pivot_longer(-date, names_to = c("var", "region"), names_pattern = "(.*)_(nz|cant)") %>%
  mutate(region = if_else(region == "nz", "NZ", "Canterbury"),
         var = factor(var, levels = c("consents_per1000", "pop_growth", "vol_yoy"),
                      labels = c("Dwelling consents, 12m per 1,000 people",
                                 "Population growth, % y/y",
                                 "Sales volume, % y/y (12m sum)")))

ggplot(by_region, aes(date, value, colour = region)) +
  geom_line(na.rm = TRUE) +
  geom_vline(xintercept = covid, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ var, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = NULL, colour = NULL, title = "Regional drivers") +
  theme_minimal() + theme(legend.position = "top")

national <- df %>%
  select(date, `2-yr mortgage rate (%)` = rate_2y, `OCR (%)` = ocr,
         `LVR >80% share of new lending (%)` = lvr_share) %>%
  pivot_longer(-date) %>%
  mutate(panel = if_else(name == "LVR >80% share of new lending (%)",
                         name, "Interest rates (%)"))

ggplot(national, aes(date, value, colour = name)) +
  geom_line(na.rm = TRUE) +
  geom_vline(xintercept = covid, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ panel, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = NULL, colour = NULL, title = "National drivers") +
  theme_minimal() + theme(legend.position = "top")
```

```{r prepost}
# Pre/post COVID means
df %>%
  filter(!is.na(pti_nz)) %>%
  mutate(period = if_else(date < covid,
                          paste0("Pre-COVID (to ", format(covid - 1, "%b %Y"), ")"),
                          paste(format(covid, "%b %Y"), "onward"))) %>%
  group_by(period) %>%
  summarise(across(c(pti_nz, pti_cant, serv_nz_2y, serv_cant_2y, rti_nz, rti_cant),
                   ~ round(mean(.x, na.rm = TRUE), 1)), .groups = "drop")
```

## What drove the change? Decomposition

Log payment burden = log price + log payment factor(rate) - log income, so changes split exactly
into price, rate and income contributions. Price-to-income has the same price and income
contributions and no rate term. Windows start and end in March, the observed income points.

```{r decomp, fig.height = 5}
k_fun <- function(rate_pct, years = 30) {          # monthly payment per $1 borrowed
  r <- rate_pct / 100 / 12
  n <- years * 12
  r / (1 - (1 + r)^(-n))
}

windows <- tibble(
  window = c("Mar 2015 to Mar 2020", "Mar 2020 to Mar 2022", "Mar 2022 to Mar 2025"),
  from   = as.Date(c("2015-03-01", "2020-03-01", "2022-03-01")),
  to     = as.Date(c("2020-03-01", "2022-03-01", "2025-03-01"))
)

anchors <- unique(c(windows$from, windows$to))   # March dates at the ends of each window

anchor <- df %>%
  filter(date %in% anchors) %>%
  transmute(date,
            NZ_price          = log(nz_price),
            NZ_rate           = log(k_fun(rate_2y)),
            NZ_income         = -log(income_nz),       # higher income lowers the burden
            Canterbury_price  = log(cant_price),
            Canterbury_rate   = log(k_fun(rate_2y)),
            Canterbury_income = -log(income_cant)) %>%
  pivot_longer(-date, names_to = c("region", "component"), names_sep = "_")

contrib <- windows %>%
  inner_join(rename(anchor, from = date, v0 = value), by = "from") %>%
  inner_join(rename(anchor, to = date, v1 = value), by = c("to", "region", "component")) %>%
  mutate(contribution = 100 * (v1 - v0),
         window    = factor(window, levels = windows$window),
         component = factor(component, levels = c("price", "rate", "income")))

# Direct change in log(outcome) between window ends, used to check the contributions sum exactly
direct_change <- function(nz_var, cant_var) {
  at <- df %>%
    select(date, NZ = {{ nz_var }}, Canterbury = {{ cant_var }}) %>%
    pivot_longer(-date, names_to = "region", values_to = "x")
  windows %>%
    inner_join(rename(at, from = date, x0 = x), by = "from") %>%
    inner_join(rename(at, to = date, x1 = x), by = c("to", "region")) %>%
    transmute(window = factor(window, levels = windows$window), region,
              direct = 100 * log(x1 / x0))
}

decomp_report <- function(cd, direct, ylab, caption) {
  totals <- cd %>%
    group_by(window, region) %>%
    summarise(total = sum(contribution), .groups = "drop")
  chk <- left_join(totals, direct, by = c("window", "region"))
  stopifnot(all(abs(chk$total - chk$direct) < 1e-8))

  tab <- cd %>%
    select(window, region, component, contribution) %>%
    pivot_wider(names_from = component, values_from = contribution) %>%
    left_join(totals, by = c("window", "region")) %>%
    mutate(across(where(is.numeric), ~ round(.x, 1))) %>%
    arrange(region, window)

  pal <- c(price = "#1b9e77", rate = "#d95f02", income = "#7570b3")
  lab <- c(price = "House price", rate = "Mortgage rate", income = "Income")
  used <- intersect(names(pal), as.character(unique(cd$component)))

  plt <- ggplot(cd, aes(window, contribution, fill = component)) +
    geom_col(width = 0.65) +
    geom_point(data = totals, aes(y = total, fill = NULL), shape = 18, size = 3,
               show.legend = FALSE) +
    geom_hline(yintercept = 0, colour = "grey40") +
    facet_wrap(~ region) +
    scale_fill_manual(values = unname(pal[used]), breaks = used, labels = unname(lab[used])) +
    labs(x = NULL, y = ylab, fill = NULL, caption = caption) +
    theme_minimal() + theme(legend.position = "top",
                            axis.text.x = element_text(angle = 20, hjust = 1))
  list(table = tab, plot = plt)
}

burden <- decomp_report(
  contrib, direct_change(serv_nz_2y, serv_cant_2y),
  "Contribution to change in payment burden (log points x100)",
  "Diamond = total change. 80% loan, 30-year term, 2-year rate, per-person income.")
burden$table
burden$plot

pti <- decomp_report(
  filter(contrib, component != "rate"), direct_change(pti_nz, pti_cant),
  "Contribution to change in price-to-income (log points x100)",
  "Diamond = total change in log price-to-income. Per-person income.")
pti$table
pti$plot
```

## COVID: actual vs pre-COVID trend

```{r counterfactual, fig.height = 8}
# Project each region's pre-COVID log-linear trend forward and compare with actual
yr_start <- seq(covid, by = "year", length.out = 5)
yr_end   <- c(yr_start[-1] - 1, max(df$date[!is.na(df$pti_nz)]))
yr_label <- sprintf("%s-%s", format(yr_start, "%b %Y"), format(yr_end, "%b %Y"))

q4 <- df %>%
  filter(!is.na(pti_nz)) %>%                 # Mar 2015 to Mar 2025
  mutate(t = row_number() / 12,              # time in years
         lpti_nz   = log(pti_nz),
         lpti_cant = log(pti_cant),
         gap       = lpti_cant - lpti_nz,
         period    = factor(c("pre", yr_label)[findInterval(date, yr_start) + 1],
                            levels = c("pre", yr_label)))

proj_dev <- function(y, data, break_date) {
  pre  <- data %>% filter(date <  break_date)
  post <- data %>% filter(date >= break_date)

  m0 <- lm(reformulate("t", y), data = pre)
  V  <- NeweyWest(m0, lag = 12, prewhite = FALSE)   # HAC covariance of the trend fit
  X  <- cbind(1, post$t)

  post <- post %>%
    mutate(actual = .data[[y]],
           proj   = as.numeric(X %*% coef(m0)),
           se     = sqrt(rowSums((X %*% V) * X)),   # uncertainty of the projection
           dev    = 100 * (actual - proj),
           lo     = 100 * (actual - proj - 1.96 * se),
           hi     = 100 * (actual - proj + 1.96 * se),
           row    = row_number())

  by_year <- post %>%
    group_by(period) %>%
    summarise(deviation = mean(dev),
              se = 100 * sqrt({xb <- colMeans(X[row, , drop = FALSE]); as.numeric(xb %*% V %*% xb)}),
              .groups = "drop") %>%
    mutate(lo = deviation - 1.96 * se, hi = deviation + 1.96 * se)

  list(monthly = select(post, date, period, dev, lo, hi),
       yearly = by_year, pre_slope = 100 * coef(m0)[["t"]])
}

outs <- c(NZ = "lpti_nz", Canterbury = "lpti_cant", `Canterbury minus NZ` = "gap")
res  <- map(outs, ~ proj_dev(.x, q4, covid))

map_dfr(res, "yearly", .id = "series") %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%
  arrange(series, period)

round(map_dbl(res, "pre_slope"), 1)       # pre-COVID trend, % per year

map_dfr(res, "monthly", .id = "series") %>%
  mutate(series = factor(series, levels = names(outs))) %>%
  ggplot(aes(date, dev)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey80") +
  geom_line(colour = "#1b9e77") +
  geom_hline(yintercept = 0, colour = "grey40") +
  facet_wrap(~ series, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Actual vs projected pre-COVID trend (log points x100)",
       caption = paste0("Log price-to-income. Band: 95% uncertainty of the projected trend (HAC). Break: ",
                        format(covid, "%b %Y"), ".")) +
  theme_minimal()
```

```{r trend-plots}
trend_plot <- function(var, region) {
  d <- df %>% filter(!is.na(.data[[var]])) %>%
    mutate(t = row_number() / 12, actual = .data[[var]])

  fit <- lm(log(actual) ~ t, data = filter(d, date < covid))     # pre-COVID trend only
  d <- d %>% mutate(predicted = exp(predict(fit, newdata = d)))

  d %>%
    select(date, Actual = actual, `Pre-COVID trend` = predicted) %>%
    pivot_longer(-date, names_to = "series", values_to = "pti") %>%
    ggplot(aes(date, pti, colour = series, linetype = series)) +
    annotate("rect", xmin = covid, xmax = max(d$date), ymin = -Inf, ymax = Inf,
             fill = "grey90", alpha = 0.6) +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = covid, colour = "grey40", linetype = "dotted") +
    scale_colour_manual(values = c(Actual = "black", `Pre-COVID trend` = "#d95f02")) +
    scale_linetype_manual(values = c(Actual = "solid", `Pre-COVID trend` = "dashed")) +
    labs(x = NULL, y = "Price-to-income (years of per-person income)",
         colour = NULL, linetype = NULL,
         title = paste0(region, ": actual vs pre-COVID trend"),
         caption = paste0("Trend fitted to pre-", format(covid, "%b %Y"),
                          " data only, then extended. Shaded: from ", format(covid, "%b %Y"), ".")) +
    theme_minimal() + theme(legend.position = "top")
}

trend_plot("pti_nz",   "NZ")
trend_plot("pti_cant", "Canterbury")
```

### Robustness of the pre-COVID trend

Does the slope, and the final-month deviation from trend, depend on the window chosen?

```{r robustness, fig.height = 5}
build_d <- function(var, region_name) {
  df %>%
    filter(!is.na(.data[[var]])) %>%
    mutate(t = row_number() / 12, actual = .data[[var]], region = region_name) %>%
    select(date, region, t, actual)
}
d <- bind_rows(build_d("pti_nz", "NZ"), build_d("pti_cant", "Canterbury"))
final_date <- max(d$date)
first_date <- as.Date("2015-03-01")
regions    <- c("NZ", "Canterbury")

# Fit the pre-COVID trend on a window (optionally dropping a block) and report slope, R2,
# and the deviation of the last observation from the projected trend (log points x100)
trend_stats <- function(region_name, from, to, drop_from, drop_to) {
  dat <- d %>% filter(region == region_name, date >= from, date < to)
  if (!is.na(drop_from)) dat <- filter(dat, !(date >= drop_from & date < drop_to))
  if (nrow(dat) < 8) return(NULL)                   # guard against degenerate windows

  fit  <- lm(log(actual) ~ t, data = dat)
  last <- filter(d, region == region_name, date == final_date)
  tibble(n_obs = nrow(dat),
         slope_pct_yr = 100 * coef(fit)[["t"]],      # t is in years
         r_squared = summary(fit)$r.squared,
         deviation_final = 100 * (log(last$actual) - predict(fit, last)))
}

no_drop <- as.Date(NA)
loyo <- tibble(block = c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20"),
               b_from = as.Date(c("2015-03-01", "2016-03-01", "2017-03-01", "2018-03-01", "2019-03-01")),
               b_to   = as.Date(c("2016-03-01", "2017-03-01", "2018-03-01", "2019-03-01", "2020-03-01")))

specs <- bind_rows(
  expand_grid(region = regions, start = as.Date(c("2015-03-01", "2016-01-01", "2017-01-01", "2018-01-01"))) %>%
    transmute(region, test = "Vary start", spec = paste("start", format(start, "%Y-%m")),
              from = start, to = covid, drop_from = no_drop, drop_to = no_drop),
  expand_grid(region = regions, end = as.Date(c("2019-01-01", "2019-07-01", "2020-01-01", "2020-03-01", "2020-04-01"))) %>%
    transmute(region, test = "Vary end", spec = paste("end", format(end, "%Y-%m")),
              from = first_date, to = end, drop_from = no_drop, drop_to = no_drop),
  expand_grid(region = regions, loyo) %>%
    transmute(region, test = "Leave-one-year-out", spec = paste("drop", block),
              from = first_date, to = covid, drop_from = b_from, drop_to = b_to)
)

robust <- specs %>%
  mutate(stats = pmap(list(region, from, to, drop_from, drop_to), trend_stats)) %>%
  unnest(stats) %>%
  select(test, region, spec, n_obs, slope_pct_yr, r_squared, deviation_final)

robust %>% mutate(across(where(is.numeric), ~ round(.x, 2)))

ggplot(robust, aes(spec, slope_pct_yr, fill = region)) +
  geom_col(position = "dodge") +
  facet_wrap(~ test, scales = "free_x") +
  geom_hline(yintercept = 0, colour = "grey40") +
  labs(x = NULL, y = "Pre-COVID trend slope (%/yr)", fill = NULL,
       title = "Robustness of pre-COVID PTI trend slope to window choice") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1), legend.position = "top")
```

## Drivers of house prices (Q1, Q2)

Outcome: log price. Income is left out as a regressor because it is the denominator of the
affordability measures. Net migration is NZ-wide for both regions (no reliable Canterbury series).
Population growth and sales volume are excluded: Canterbury population is annual-only (interpolated,
so near-collinear with time) and volume is jointly determined with price. HAC (Newey-West) standard
errors throughout.

```{r models}
# Stack the two regions under common variable names so one formula serves both
model_data <- bind_rows(
  df %>% transmute(date, region = "NZ", price = nz_price, consents = consents_per1000_nz,
                   rate_2y, lvr_share, net_mig_nz_k, unemp, bright_line, inventory = invent_nz),
  df %>% transmute(date, region = "Canterbury", price = cant_price, consents = consents_per1000_cant,
                   rate_2y, lvr_share, net_mig_nz_k, unemp, bright_line, inventory = invent_cant)
) %>%
  mutate(post_covid = as.integer(date >= covid))

if (common_window) {   # full-sample models on the months where both regions have every variable
  complete_dates <- model_data %>%
    drop_na() %>% count(date) %>% filter(n == 2) %>% pull(date)
  model_data <- filter(model_data, date %in% complete_dates)
}

f_pre  <- log(price) ~ rate_2y + lvr_share + net_mig_nz_k + consents + unemp + bright_line + inventory
# Interactions on rate and migration let those slopes change after COVID; the interaction
# p-value tests a break, and post-COVID slope = main effect + interaction
f_full <- update(f_pre, . ~ . + post_covid + rate_2y:post_covid + net_mig_nz_k:post_covid + inventory:post_covid)

fit_one <- function(region_name, sample) {
  dat <- filter(model_data, region == region_name)
  # do.call embeds the data in the model call, so waldtest()/car::vif() can re-evaluate it
  if (sample == "pre") do.call(lm, list(f_pre, data = filter(dat, date < covid)))
  else do.call(lm, list(f_full, data = dat))
}

fits <- list(
  "NZ (pre-COVID)"           = fit_one("NZ", "pre"),
  "Canterbury (pre-COVID)"   = fit_one("Canterbury", "pre"),
  "NZ (full sample)"         = fit_one("NZ", "full"),
  "Canterbury (full sample)" = fit_one("Canterbury", "full")
)

hac_tidy <- function(fit, model_name) {
  ct <- coeftest(fit, vcov = NeweyWest(fit, prewhite = FALSE, adjust = TRUE))
  tibble(model = model_name, term = rownames(ct),
         estimate = ct[, 1], std.error = ct[, 2], statistic = ct[, 3], p.value = ct[, 4],
         r.squared = summary(fit)$r.squared, n = nobs(fit))
}

sig_stars <- function(p) {
  case_when(p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", p < 0.1 ~ ".", TRUE ~ "")
}

results <- imap_dfr(fits, hac_tidy) %>%
  mutate(model = factor(model, levels = names(fits)),
         display = paste0(sprintf("%.4f", estimate), sig_stars(p.value)))
```

```{r model-tables, results = "asis"}
for (m in names(fits)) {
  r <- filter(results, model == m)
  cat(sprintf("\n#### %s (R-squared = %.3f, n = %d)\n\n", m, r$r.squared[1], r$n[1]))
  r %>%
    transmute(Term = term, Estimate = display,
              `Std.Error` = sprintf("%.4f", std.error),
              `t-stat` = sprintf("%.2f", statistic),
              `p-value` = sprintf("%.3f", p.value)) %>%
    knitr::kable(align = "lrrrr") %>%
    cat(sep = "\n")
}
cat("\nSignificance: *** p<0.001, ** p<0.01, * p<0.05, . p<0.1\n")
```

```{r model-diagnostics}
# Joint HAC Wald test: are all slope coefficients zero?
imap_dfr(fits, function(fit, m) {
  w <- waldtest(fit, vcov = NeweyWest(fit, prewhite = FALSE, adjust = TRUE))
  tibble(model = m, F_stat = round(w$F[2], 3), p_value = round(w$`Pr(>F)`[2], 4))
})

# VIF. Main-effect/interaction pairs in the full-sample models are collinear by construction
imap_dfr(fits, ~ tibble(model = .y, term = names(car::vif(.x)), VIF = round(car::vif(.x), 2))) %>%
  pivot_wider(names_from = model, values_from = VIF) %>%
  arrange(term)

# Q2: NZ vs Canterbury, all terms side by side
results %>%
  filter(term != "(Intercept)") %>%
  select(model, term, display) %>%
  pivot_wider(names_from = model, values_from = display, values_fill = "")
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

covid <- as.Date("2020-03-01")

make_stock_flow_panel <- function(data, region_name, consents_col, invent_col) {
  d <- data %>%
    transmute(date,
              consents  = .data[[consents_col]],
              inventory = .data[[invent_col]]) %>%
    filter(!is.na(consents) | !is.na(inventory))

  rng_consents  <- range(d$consents,  na.rm = TRUE)
  rng_inventory <- range(d$inventory, na.rm = TRUE)
  scale_factor  <- diff(rng_consents) / diff(rng_inventory)
  offset        <- rng_consents[1] - rng_inventory[1] * scale_factor

  d <- d %>% mutate(inventory_scaled = inventory * scale_factor + offset)

  ggplot(d, aes(date)) +
    geom_line(aes(y = consents, colour = "Consents per 1,000 (flow)"), linewidth = 0.7) +
    geom_line(aes(y = inventory_scaled, colour = "Weeks of inventory (stock)"), linewidth = 0.7) +
    geom_vline(xintercept = covid, linetype = "dashed", colour = "grey40") +
    scale_y_continuous(
      name = "Consents per 1,000 people",
      sec.axis = sec_axis(~ (. - offset) / scale_factor, name = "Weeks of inventory")
    ) +
    scale_colour_manual(values = c("Consents per 1,000 (flow)" = "#1b9e77",
                                    "Weeks of inventory (stock)" = "#d95f02")) +
    labs(x = NULL, colour = NULL, title = region_name) +
    theme_minimal() +
    theme(legend.position = "top")
}

p_nz   <- make_stock_flow_panel(df, "NZ",         "consents_per1000_nz",   "invent_nz")
p_cant <- make_stock_flow_panel(df, "Canterbury", "consents_per1000_cant", "invent_cant")

p_nz
p_cant

```

```{r}
## ---- chunk 15: coefficient plot for the driver regressions ----------------
# Charts the regression results directly, rather than just tabulating them.
# A dot-and-whisker (coefficient) plot is used, not a scatter of the raw
# data, because these are MULTIPLE regression models (6-11 regressors each)
# -- there's no single x/y pair to scatter. A coefficient plot is the
# standard way to visualise multi-regressor OLS output: each term gets a
# point estimate + 95% CI whisker, faceted/grouped so you can compare the
# same term's estimate and precision across all 4 models at once.
#
# This directly visualises everything already established in discussion:
# sign flips (rate_2y, inventory), which terms are consistently robust
# (unemp) vs unstable across specifications, and the NZ vs Canterbury
# divergence -- in one chart instead of four separate tables.
#
# Reuses `results` (built by hac_tidy() + imap_dfr(fits, hac_tidy) in the
# models chunk) directly -- no need to refit anything. CI built from the
# already-computed HAC std.error (95% CI = estimate +/- 1.96*std.error),
# consistent with the Newey-West SEs used throughout the rest of the report.

library(dplyr)
library(ggplot2)
library(stringr)

## ---- build CI + tidy labels from the existing `results` tibble -----------
coef_plot_data <- results %>%
  filter(term != "(Intercept)") %>%     # intercept isn't economically interesting here
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    sample    = if_else(str_detect(model, "pre-COVID"), "Pre-COVID", "Full sample"),
    region    = if_else(str_detect(model, "Canterbury"), "Canterbury", "NZ"),
    sample    = factor(sample, levels = c("Pre-COVID", "Full sample"))
  )

## ---- order terms sensibly (main effects first, then their interactions) --
term_order <- c("rate_2y", "rate_2y:post_covid",
                 "lvr_share",
                 "net_mig_nz_k", "net_mig_nz_k:post_covid",
                 "consents",
                 "inventory", "inventory:post_covid",
                 "unemp", "bright_line", "post_covid")
coef_plot_data <- coef_plot_data %>%
  mutate(term = factor(term, levels = rev(term_order)))

## ---- plot: one point + whisker per term, coloured by region, faceted by sample
ggplot(coef_plot_data, aes(x = estimate, y = term, colour = region)) +
  geom_vline(xintercept = 0, colour = "grey50", linetype = "dashed") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                   position = position_dodge(width = 0.5), size = 0.4) +
  facet_wrap(~ sample, scales = "free_x") +
  scale_colour_manual(values = c("NZ" = "#1b9e77", "Canterbury" = "#d95f02")) +
  labs(x = "Coefficient estimate (95% CI, HAC/Newey-West SEs)", y = NULL,
       colour = NULL,
       title = "Driver regression coefficients: NZ vs Canterbury, pre-COVID vs full sample",
       caption = "Dashed line = zero effect. Intercept omitted. Whiskers crossing zero indicate a non-significant term at 5%.") +
  theme_minimal() +
  theme(legend.position = "top", strip.text = element_text(face = "bold"))

## ---- optional companion: actual vs fitted price, per region --------------
# Useful alongside the coefficient plot to SHOW model fit quality, not just
# state the R-squared -- particularly for Canterbury, where R-squared is
# much lower (0.58-0.61 pre-COVID) than NZ's (0.92+). Seeing the fitted
# line visibly track the actual series more loosely for Canterbury makes
# that gap concrete rather than just a number in a table.

library(broom)
library(purrr)
library(tidyr)

fitted_data <- imap_dfr(fits, function(fit, model_name) {
  broom::augment(fit) %>%
    mutate(model = model_name,
           region = if_else(str_detect(model_name, "Canterbury"), "Canterbury", "NZ"),
           sample = if_else(str_detect(model_name, "pre-COVID"), "Pre-COVID", "Full sample"))
})

# augment() doesn't carry `date` through automatically -- re-attach it by
# row order within each model's own data (safe since no rows are reordered
# by lm()/augment()).
fitted_data <- fitted_data %>%
  group_by(model) %>%
  mutate(date = model_data %>%
           filter(region == first(region),
                  if (first(sample) == "Pre-COVID") date < covid else TRUE) %>%
           pull(date)) %>%
  ungroup()

ggplot(fitted_data, aes(date)) +
  geom_line(aes(y = `log(price)`, colour = "Actual"), linewidth = 0.6) +
  geom_line(aes(y = .fitted, colour = "Fitted"), linewidth = 0.6, linetype = "dashed") +
  facet_grid(region ~ sample, scales = "free") +
  scale_colour_manual(values = c("Actual" = "black", "Fitted" = "#1b9e77")) +
  labs(x = NULL, y = "log(price)", colour = NULL,
       title = "Actual vs fitted log(price): model fit by region and sample") +
  theme_minimal() + theme(legend.position = "top")

```

```{r}
f_split <- log(price) ~ rate_2y + lvr_share + net_mig_nz_k + consents + unemp + bright_line

split_fit <- function(region_name, period) {
  dat <- model_data %>%
    filter(region == region_name, if (period == "pre") date < covid else date >= covid) %>%
    drop_na(price, rate_2y, lvr_share, net_mig_nz_k, consents, unemp, bright_line)
  do.call(lm, list(f_split, data = dat))
}

grid <- expand_grid(region = c("NZ", "Canterbury"), period = c("pre", "post"))
split_fits <- pmap(grid, ~ split_fit(..1, ..2)) %>%
  set_names(paste0(grid$region, " (", grid$period, "-COVID)"))

results_split <- imap_dfr(split_fits, hac_tidy) %>%
  mutate(display = paste0(sprintf("%.4f", estimate), sig_stars(p.value)))

# Coefficients side by side: pre vs post
terms_order <- c("rate_2y", "lvr_share", "net_mig_nz_k", "consents", "unemp", "bright_line")
results_split %>%
  filter(term != "(Intercept)") %>%
  separate(model, into = c("region", "period"), sep = " \\(") %>%
  mutate(period = str_remove(period, "-COVID\\)")) %>%
  select(region, term, period, display) %>%
  pivot_wider(names_from = period, values_from = display) %>%
  arrange(region, match(term, terms_order)) %>%
  select(region, term, pre, post)

# Fit summary
results_split %>% distinct(model, r.squared, n) %>% mutate(r.squared = round(r.squared, 3))

# Did the relationships change? HAC Wald test that every slope is equal pre vs post
hac <- function(x) NeweyWest(x, prewhite = FALSE, adjust = TRUE)
map_dfr(c("NZ", "Canterbury"), function(r) {
  dat <- model_data %>% filter(region == r) %>%
    drop_na(price, rate_2y, lvr_share, net_mig_nz_k, consents, unemp, bright_line)
  pooled <- do.call(lm, list(f_split, data = dat))
  broken <- do.call(lm, list(update(f_split, . ~ . + post_covid * (rate_2y + lvr_share + net_mig_nz_k +
                                                                      consents + unemp + bright_line)), data = dat))
  w <- waldtest(broken, pooled, vcov = hac)
  tibble(region = r, F_stat = round(w$F[2], 2), p_value = round(w$`Pr(>F)`[2], 4))
})
```



```{r}
panel_levels <- c("NZ (pre-COVID)", "NZ (post-COVID)", "Canterbury (pre-COVID)", "Canterbury (post-COVID)")

fit_df <- pmap_dfr(grid, function(region, period) {
  reg <- region; per <- period
  fit <- split_fits[[paste0(reg, " (", per, "-COVID)")]]
  dat <- model_data %>%
    filter(region == reg, if (per == "pre") date < covid else date >= covid) %>%
    drop_na(price, rate_2y, lvr_share, net_mig_nz_k, consents, unemp, bright_line)

  tibble(date = dat$date, Actual = dat$price / 1000, Fitted = exp(predict(fit, newdata = dat)) / 1000,
         panel = sprintf("%s, R\u00b2 = %.2f", paste0(reg, " (", per, "-COVID)"), summary(fit)$r.squared),
         order = match(paste0(reg, " (", per, "-COVID)"), panel_levels))
}) %>%
  mutate(panel = fct_reorder(panel, order)) %>%
  pivot_longer(c(Actual, Fitted), names_to = "series", values_to = "price_k")

ggplot(fit_df, aes(date, price_k, colour = series, linetype = series)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ panel, scales = "free", ncol = 2) +
  scale_colour_manual(values = c(Actual = "black", Fitted = "#d95f02")) +
  scale_linetype_manual(values = c(Actual = "solid", Fitted = "dashed")) +
  labs(x = NULL, y = "Median sale price ($000)", colour = NULL, linetype = NULL,
       title = "Actual vs fitted median sale price, split pre- and post-COVID models",
       caption = "In-sample fit: each model is fitted and evaluated on its own period, so this shows fit, not prediction.") +
  theme_minimal() + theme(legend.position = "top")
```

```{r}
f_split <- log(price) ~ rate_2y + lvr_share + net_mig_nz_k + consents + unemp + bright_line
vars    <- c("price", "rate_2y", "lvr_share", "net_mig_nz_k", "consents", "unemp", "bright_line")

# Fit on earlier months, predict a held-out final 12 months using the actual driver values,
# and compare with a no-change forecast (last training-period price)
holdout_test <- function(region_name, label, train_from, train_to, test_to) {
  d  <- model_data %>% filter(region == region_name) %>% drop_na(all_of(vars))
  tr <- filter(d, date >= train_from, date <= train_to)
  te <- filter(d, date >  train_to,   date <= test_to)

  pred   <- predict(lm(f_split, data = tr), newdata = te)
  actual <- log(te$price)
  naive  <- log(tail(tr$price, 1))

  tibble(region = region_name, model = label, n_train = nrow(tr), n_test = nrow(te),
         model_error = 100 * sqrt(mean((actual - pred)^2)),   # RMSE, log points x100
         nochange_error = 100 * sqrt(mean((actual - naive)^2)),
         bias = 100 * mean(actual - pred))                    # >0: model underpredicts
}

pre_test_end  <- covid %m-% months(1)                  # last pre-COVID month
post_test_end <- max(model_data$date)                  # last month in the data

map_dfr(c("NZ", "Canterbury"), function(r) {
  bind_rows(
    holdout_test(r, "Pre-COVID model",  min(model_data$date), pre_test_end %m-% months(12), pre_test_end),
    holdout_test(r, "Post-COVID model", covid,                post_test_end %m-% months(12), post_test_end))
}) %>%
  mutate(relative_to_nochange = model_error / nochange_error,
         across(where(is.numeric), ~ round(.x, 1)))
```

