---
title: "DATA309 Housing Affordability — Statistical Analysis"
author: "Mohammad Zaid Khan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(tidyverse)
library(readxl)
library(broom)
library(lmtest)
library(sandwich)
library(car)
library(knitr)

# ------------------------------------------------------------
# Helper functions
# ------------------------------------------------------------

first_existing <- function(files) {

  found <- files[file.exists(files)]

  if(length(found) == 0) {
    stop(
      paste(
        "Could not find any of these files:",
        paste(files, collapse = ", ")
      )
    )
  }

  found[1]
}


nw_tidy <- function(model, model_name, lag_length = 12) {

  test <- coeftest(
    model,
    vcov. = NeweyWest(
      model,
      lag = lag_length,
      prewhite = FALSE,
      adjust = TRUE
    )
  )

  data.frame(
    Model = model_name,
    term = rownames(test),
    estimate = test[, 1],
    std.error = test[, 2],
    statistic = test[, 3],
    p.value = test[, 4],
    row.names = NULL
  )
}


first_non_missing <- function(x) {

  x <- x[!is.na(x)]

  if(length(x) == 0) {
    return(NA_real_)
  }

  x[1]
}
```

# 1. Analysis Overview

This analysis investigates housing affordability and housing-market
conditions in Canterbury.

The analysis focuses on three questions:

1. How does housing affordability in Canterbury compare with New Zealand?
2. What economic and housing-market factors are associated with
   Canterbury house-price growth?
3. Have household incomes and rents kept pace with Canterbury house
   prices?

The analysis uses descriptive statistics, graphical exploratory analysis,
paired tests, regression modelling, lagged explanatory variables and
time-series diagnostics.

Because the data are observational, regression coefficients are
interpreted as **associations rather than causal effects**.

---

# 2. Data Import

All datasets should be stored in the same project folder as this R
Markdown file.

```{r import}
affordability_file <- first_existing(
  c(
    "house-value-to-income-multiple-and-mortgage-payment-proportion-of-income (1).csv",
    "house-value-to-income-multiple-and-mortgage-payment-proportion-of-income.csv"
  )
)

time_series_file <- first_existing(
  c(
    "Time_series_data.xlsx",
    "Time_series_data(1).xlsx",
    "Time_series_data(2).xlsx",
    "Time_series_data (1).xlsx",
    "Time_series_data (2).xlsx"
  )
)

affordability <- read.csv(
  affordability_file,
  stringsAsFactors = FALSE
)

# Standardise CSV column names
names(affordability) <- make.names(
  names(affordability)
)

ts_raw <- read_excel(
  time_series_file
)

cat(
  "Affordability file:",
  affordability_file,
  "\n"
)

cat(
  "Time-series file:",
  time_series_file,
  "\n"
)
```

# 3. Analysis 1 — Canterbury vs New Zealand Affordability

## 3.1 Prepare affordability data

```{r analysis1-prepare}
affordability_compare <- affordability %>%
  transmute(
    Year =
      Year,

    Canterbury_ValueIncome =
      Canterbury.Region.House.value.to.income,

    NZ_ValueIncome =
      New.Zealand.House.value.to.income,

    Canterbury_Mortgage =
      Canterbury.Region.Mortgage.payment.proportion.of.income.,

    NZ_Mortgage =
      New.Zealand.Mortgage.payment.proportion.of.income.
  ) %>%
  mutate(
    ValueIncome_Gap =
      Canterbury_ValueIncome -
      NZ_ValueIncome,

    Mortgage_Gap =
      Canterbury_Mortgage -
      NZ_Mortgage
  ) %>%
  arrange(Year)

kable(
  affordability_compare,
  digits = 2,
  col.names = c(
    "Year",
    "Canterbury Value-to-Income",
    "NZ Value-to-Income",
    "Canterbury Mortgage Burden (%)",
    "NZ Mortgage Burden (%)",
    "Value-to-Income Gap",
    "Mortgage Gap (pp)"
  ),
  caption =
    "Canterbury and New Zealand Affordability Measures"
)
```

## 3.2 Average affordability

```{r analysis1-summary}
comparison_summary <- affordability_compare %>%
  summarise(
    Canterbury_Mean_ValueIncome =
      mean(
        Canterbury_ValueIncome,
        na.rm = TRUE
      ),

    NZ_Mean_ValueIncome =
      mean(
        NZ_ValueIncome,
        na.rm = TRUE
      ),

    Mean_ValueIncome_Gap =
      mean(
        ValueIncome_Gap,
        na.rm = TRUE
      ),

    Canterbury_Mean_Mortgage =
      mean(
        Canterbury_Mortgage,
        na.rm = TRUE
      ),

    NZ_Mean_Mortgage =
      mean(
        NZ_Mortgage,
        na.rm = TRUE
      ),

    Mean_Mortgage_Gap =
      mean(
        Mortgage_Gap,
        na.rm = TRUE
      )
  )

kable(
  comparison_summary,
  digits = 2,
  col.names = c(
    "Cant. HVI",
    "NZ HVI",
    "HVI Gap",
    "Cant. Mortgage (%)",
    "NZ Mortgage (%)",
    "Mortgage Gap (pp)"
  ),
  caption =
    "Average Affordability: Canterbury vs New Zealand"
)
```

### Interpretation

Across the available period, Canterbury's average house value-to-income
multiple was approximately **4.99**, compared with approximately
**5.75** nationally.

Canterbury's average mortgage-payment burden was approximately
**31.37%**, compared with approximately **35.89%** nationally.

This suggests that Canterbury has generally been more affordable than
New Zealand overall.

---

## 3.3 House value-to-income comparison

```{r analysis1-valueincome-plot}
value_income_long <- affordability_compare %>%
  select(
    Year,
    Canterbury =
      Canterbury_ValueIncome,
    `New Zealand` =
      NZ_ValueIncome
  ) %>%
  pivot_longer(
    -Year,
    names_to = "Region",
    values_to = "ValueIncome"
  )

ggplot(
  value_income_long,
  aes(
    x = Year,
    y = ValueIncome,
    colour = Region
  )
) +
  geom_line(
    linewidth = 1.1
  ) +
  geom_point(
    size = 2
  ) +
  labs(
    title =
      "House Value-to-Income: Canterbury vs New Zealand",
    subtitle =
      "2005–2026",
    x = "Year",
    y =
      "House value-to-income multiple",
    colour = NULL
  ) +
  theme_minimal(
    base_size = 13
  )
```

---

## 3.4 Mortgage-payment burden comparison

```{r analysis1-mortgage-plot}
mortgage_long <- affordability_compare %>%
  select(
    Year,
    Canterbury =
      Canterbury_Mortgage,
    `New Zealand` =
      NZ_Mortgage
  ) %>%
  pivot_longer(
    -Year,
    names_to = "Region",
    values_to = "MortgageBurden"
  )

ggplot(
  mortgage_long,
  aes(
    x = Year,
    y = MortgageBurden,
    colour = Region
  )
) +
  geom_line(
    linewidth = 1.1
  ) +
  geom_point(
    size = 2
  ) +
  labs(
    title =
      "Mortgage Payment Burden: Canterbury vs New Zealand",
    subtitle =
      "2005–2026",
    x = "Year",
    y =
      "Mortgage payment proportion of income (%)",
    colour = NULL
  ) +
  theme_minimal(
    base_size = 13
  )
```

---

## 3.5 Affordability gap

```{r analysis1-gap}
ggplot(
  affordability_compare,
  aes(
    x = Year,
    y = ValueIncome_Gap
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    colour = "grey40"
  ) +
  geom_line(
    colour = "#D55E00",
    linewidth = 1.1
  ) +
  geom_point(
    colour = "#D55E00",
    size = 2
  ) +
  labs(
    title =
      "Canterbury's House Value-to-Income Gap Relative to NZ",
    subtitle =
      "Negative values indicate lower ratios in Canterbury",
    x = "Year",
    y =
      "Canterbury minus New Zealand"
  ) +
  theme_minimal(
    base_size = 13
  )
```

---

## 3.6 Paired statistical tests

Because the Canterbury and New Zealand observations refer to the same
years, paired tests are used.

```{r analysis1-tests}
value_test <- t.test(
  affordability_compare$Canterbury_ValueIncome,
  affordability_compare$NZ_ValueIncome,
  paired = TRUE
)

mortgage_test <- t.test(
  affordability_compare$Canterbury_Mortgage,
  affordability_compare$NZ_Mortgage,
  paired = TRUE
)

value_test

mortgage_test
```

### Finding

The mean Canterbury house value-to-income multiple was approximately
**0.76 lower than New Zealand**.

This difference was statistically significant:

**p < 0.001**.

Canterbury's mortgage-payment burden was approximately **4.52 percentage
points lower than New Zealand**, also with:

**p < 0.001**.

Therefore, Canterbury has historically been significantly more
affordable than New Zealand overall on both measures.

However, affordability still deteriorated considerably during the early
2020s.

---

# 4. Analysis 2 — What Is Associated with Canterbury House-Price Growth?

## 4.1 Prepare monthly housing-market data

```{r analysis2-prepare}
analysis2 <- ts_raw %>%
  transmute(
    Date =
      as.Date(Date),

    Canterbury_Price =
      as.numeric(
        `Canterbury Median sale price`
      ),

    NZ_Price =
      as.numeric(
        `NZ Median sale price`
      ),

    Mortgage_1Y =
      as.numeric(
        `1 Year Mortgage Rate`
      ),

    Net_Migration =
      as.numeric(
        `Net migration`
      ),

    BrightLine =
      as.numeric(
        `Bright Line Test Period`
      ),

    Canterbury_Sales =
      as.numeric(
        `Volume Sold Canterbury`
      ),

    LVR_Total =
      as.numeric(
        `LVR Total New Commitments (NZDm)`
      ),

    Canterbury_Consents =
      as.numeric(
        `Canterbury Dwelling Unit Consents`
      ),

    Canterbury_Rent =
      as.numeric(
        `Canterbury Median Rent`
      ),

    NZ_Rent =
      as.numeric(
        `Median Rent`
      ),

    Canterbury_Income =
      as.numeric(
        `Mean Income Canterbury`
      ),

    NZ_Income =
      as.numeric(
        `Mean Income`
      )
  ) %>%
  arrange(Date)
```

---

## 4.2 Create growth rates and lagged variables

Year-on-year changes are used to reduce the risk of misleading
relationships caused purely by common upward trends.

```{r analysis2-transform}
analysis2 <- analysis2 %>%
  mutate(
    Year =
      as.integer(
        format(
          Date,
          "%Y"
        )
      ),

    Canterbury_Price_YoY =
      100 * (
        Canterbury_Price /
          lag(
            Canterbury_Price,
            12
          ) - 1
      ),

    NZ_Price_YoY =
      100 * (
        NZ_Price /
          lag(
            NZ_Price,
            12
          ) - 1
      ),

    Canterbury_Sales_YoY =
      100 * (
        Canterbury_Sales /
          lag(
            Canterbury_Sales,
            12
          ) - 1
      ),

    LVR_Total_YoY =
      100 * (
        LVR_Total /
          lag(
            LVR_Total,
            12
          ) - 1
      ),

    Canterbury_Consents_YoY =
      100 * (
        Canterbury_Consents /
          lag(
            Canterbury_Consents,
            12
          ) - 1
      ),

    Migration_Change12 =
      Net_Migration -
      lag(
        Net_Migration,
        12
      ),

    Mortgage1Y_Lag6 =
      lag(
        Mortgage_1Y,
        6
      ),

    Migration_Lag6 =
      lag(
        Migration_Change12,
        6
      ),

    LVRGrowth_Lag6 =
      lag(
        LVR_Total_YoY,
        6
      ),

    Consents_Lag12 =
      lag(
        Canterbury_Consents_YoY,
        12
      ),

    SalesGrowth_Lag1 =
      lag(
        Canterbury_Sales_YoY,
        1
      )
  )
```

---

## 4.3 Canterbury vs national house-price growth

```{r analysis2-national-plot}
ggplot(
  analysis2,
  aes(
    x = NZ_Price_YoY,
    y = Canterbury_Price_YoY
  )
) +
  geom_point(
    alpha = 0.65,
    colour = "#D55E00"
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    colour = "#0072B2"
  ) +
  labs(
    title =
      "Canterbury House-Price Growth vs New Zealand",
    subtitle =
      "Year-on-year median sale-price growth",
    x =
      "New Zealand house-price growth (%)",
    y =
      "Canterbury house-price growth (%)"
  ) +
  theme_minimal(
    base_size = 13
  )
```

---

## 4.4 Key regression models

Only the models that contribute directly to the final interpretation are
retained.

```{r analysis2-models}
model_nz <- lm(
  Canterbury_Price_YoY ~
    NZ_Price_YoY,
  data = analysis2
)

model_rate <- lm(
  Canterbury_Price_YoY ~
    Mortgage1Y_Lag6,
  data = analysis2
)

model_credit <- lm(
  Canterbury_Price_YoY ~
    LVRGrowth_Lag6,
  data = analysis2
)

main_model <- lm(
  Canterbury_Price_YoY ~
    NZ_Price_YoY +
    Mortgage1Y_Lag6 +
    Migration_Lag6 +
    LVRGrowth_Lag6 +
    Consents_Lag12 +
    SalesGrowth_Lag1 +
    BrightLine,
  data = analysis2
)
```

---

## 4.5 Model comparison

```{r analysis2-comparison}
analysis2_models <- list(
  "NZ housing market" =
    model_nz,

  "Mortgage rate" =
    model_rate,

  "Credit growth" =
    model_credit,

  "Combined model" =
    main_model
)

model_comparison <- purrr::imap_dfr(
  analysis2_models,
  ~ broom::glance(.x) %>%
    mutate(
      Model = .y
    )
) %>%
  select(
    Model,
    r.squared,
    adj.r.squared,
    statistic,
    p.value,
    nobs,
    AIC,
    BIC
  )

kable(
  model_comparison,
  digits = 3,
  col.names = c(
    "Model",
    "R²",
    "Adjusted R²",
    "F-statistic",
    "p-value",
    "Observations",
    "AIC",
    "BIC"
  ),
  caption =
    "Canterbury House-Price Model Comparison"
)
```

---

## 4.6 Newey-West robust inference

The housing time series contain serial correlation.

Therefore, Newey-West robust standard errors are used for the final
statistical inference.

```{r analysis2-newey-west}
analysis2_robust <- bind_rows(

  nw_tidy(
    model_nz,
    "NZ housing market"
  ),

  nw_tidy(
    model_rate,
    "Mortgage rate"
  ),

  nw_tidy(
    model_credit,
    "Credit growth"
  ),

  nw_tidy(
    main_model,
    "Combined model"
  )
) %>%
  mutate(
    term = dplyr::recode(
      term,
      "(Intercept)" =
        "Intercept",

      "NZ_Price_YoY" =
        "NZ house-price growth",

      "Mortgage1Y_Lag6" =
        "1-year mortgage rate (6m lag)",

      "Migration_Lag6" =
        "Net migration change (6m lag)",

      "LVRGrowth_Lag6" =
        "Credit growth (6m lag)",

      "Consents_Lag12" =
        "Dwelling consents growth (12m lag)",

      "SalesGrowth_Lag1" =
        "Sales growth (1m lag)",

      "BrightLine" =
        "Bright-Line period"
    )
  )

kable(
  analysis2_robust,
  digits = 4,
  col.names = c(
    "Model",
    "Variable",
    "Estimate",
    "Robust SE",
    "t-statistic",
    "p-value"
  ),
  caption =
    "Newey-West Robust Results for Canterbury House-Price Growth"
)
```

### Main results

The strongest and most consistent result was the relationship between
Canterbury and the wider New Zealand housing market.

In the combined model, New Zealand house-price growth had an estimated
coefficient of approximately:

**0.706**

with:

**p < 0.001**.

This means that, holding the other included variables constant, a
one percentage-point increase in annual New Zealand house-price growth
was associated with approximately **0.71 percentage points higher
Canterbury house-price growth**.

The combined model produced approximately:

**R² = 0.823**

and:

**Adjusted R² = 0.811**.

Mortgage rates showed a significant negative relationship when analysed
alone.

The standalone six-month-lagged mortgage-rate coefficient was
approximately:

**-4.10**

with:

**p = 0.026**.

Credit growth also displayed a significant positive standalone
association.

However, these relationships weakened after national housing conditions
and the other variables were controlled for.

This suggests that Canterbury's housing market is strongly connected to
the broader New Zealand housing cycle.

---

## 4.7 Final diagnostics

Only the diagnostics required to justify the final regression inference
are retained.

```{r analysis2-diagnostics}
bg_main <- bgtest(
  main_model,
  order = 12
)

vif_main <- car::vif(
  main_model
)

bp_main <- bptest(
  main_model
)

bg_main

vif_main

bp_main
```

### Diagnostic interpretation

The Breusch-Godfrey test identified significant residual serial
correlation:

**p < 0.001**.

This supports the use of Newey-West robust standard errors.

The largest VIF values were approximately **4.6**, indicating moderate
but not severe multicollinearity.

The Breusch-Pagan test produced:

**p = 0.225**.

Therefore, there was no significant evidence of heteroskedasticity in
the combined model.

---

# 5. Analysis 3 — House Prices, Rents and Income

## 5.1 Purpose

This analysis examines whether household incomes and rents kept pace
with Canterbury house prices.

It connects housing-market movements more directly to household
affordability.

---

## 5.2 Monthly data

```{r analysis3-prepare}
analysis3_monthly <- analysis2 %>%
  select(
    Date,
    Year,
    Canterbury_Price,
    NZ_Price,
    Canterbury_Rent,
    NZ_Rent,
    Canterbury_Income,
    NZ_Income
  ) %>%
  mutate(
    Canterbury_Price_YoY =
      100 * (
        Canterbury_Price /
          lag(
            Canterbury_Price,
            12
          ) - 1
      ),

    Canterbury_Rent_YoY =
      100 * (
        Canterbury_Rent /
          lag(
            Canterbury_Rent,
            12
          ) - 1
      ),

    PriceGrowth_Lag12 =
      lag(
        Canterbury_Price_YoY,
        12
      ),

    Period =
      case_when(
        Year <= 2019 ~
          "2015-2019",

        Year <= 2022 ~
          "2020-2022",

        TRUE ~
          "2023 onwards"
      ),

    Period =
      factor(
        Period,
        levels = c(
          "2015-2019",
          "2020-2022",
          "2023 onwards"
        )
      )
  )
```

---

## 5.3 Annual prices, rents and income

Income is available annually rather than monthly.

Missing monthly values are not artificially interpolated.

```{r analysis3-annual}
analysis3_annual <- analysis3_monthly %>%
  group_by(
    Year
  ) %>%
  summarise(
    Canterbury_Price =
      median(
        Canterbury_Price,
        na.rm = TRUE
      ),

    NZ_Price =
      median(
        NZ_Price,
        na.rm = TRUE
      ),

    Canterbury_Rent =
      median(
        Canterbury_Rent,
        na.rm = TRUE
      ),

    NZ_Rent =
      median(
        NZ_Rent,
        na.rm = TRUE
      ),

    Canterbury_Income =
      first_non_missing(
        Canterbury_Income
      ),

    NZ_Income =
      first_non_missing(
        NZ_Income
      ),

    .groups = "drop"
  )
```

---

## 5.4 Indexed house prices, rents and income

```{r analysis3-index}
base_year <- min(
  analysis3_annual$Year,
  na.rm = TRUE
)

index_base <- analysis3_annual %>%
  filter(
    Year == base_year
  )

analysis3_index <- analysis3_annual %>%
  mutate(
    House_Prices =
      100 *
      Canterbury_Price /
      index_base$Canterbury_Price,

    Rents =
      100 *
      Canterbury_Rent /
      index_base$Canterbury_Rent,

    Income =
      100 *
      Canterbury_Income /
      index_base$Canterbury_Income
  ) %>%
  select(
    Year,
    House_Prices,
    Rents,
    Income
  ) %>%
  pivot_longer(
    -Year,
    names_to = "Measure",
    values_to = "Index"
  )

ggplot(
  analysis3_index,
  aes(
    x = Year,
    y = Index,
    colour = Measure
  )
) +
  geom_line(
    linewidth = 1.1,
    na.rm = TRUE
  ) +
  geom_point(
    size = 2,
    na.rm = TRUE
  ) +
  labs(
    title =
      "Canterbury House Prices, Rents and Income",
    subtitle =
      "Indexed comparison: first available observation = 100",
    x = "Year",
    y = "Index",
    colour = NULL
  ) +
  theme_minimal(
    base_size = 13
  ) +
  theme(
    legend.position = "bottom"
  )
```

---

## 5.5 House-price-to-income proxy

This is a separate measure from the official house value-to-income
multiple used in Analysis 1.

```{r analysis3-price-income}
price_income_table <- analysis3_annual %>%
  filter(
    !is.na(
      Canterbury_Income
    ),
    !is.na(
      NZ_Income
    )
  ) %>%
  mutate(
    Canterbury_PriceIncome =
      Canterbury_Price /
      Canterbury_Income,

    NZ_PriceIncome =
      NZ_Price /
      NZ_Income
  )

kable(
  price_income_table %>%
    select(
      Year,
      Canterbury_Price,
      Canterbury_Income,
      Canterbury_PriceIncome,
      NZ_Price,
      NZ_Income,
      NZ_PriceIncome
    ),
  digits = 2,
  col.names = c(
    "Year",
    "Canterbury House Price ($)",
    "Canterbury Income ($)",
    "Canterbury Price-to-Income",
    "NZ House Price ($)",
    "NZ Income ($)",
    "NZ Price-to-Income"
  ),
  caption =
    "House-Price-to-Income Proxy"
)
```

---

## 5.6 Canterbury price-to-income trend

```{r analysis3-price-income-plot}
ggplot(
  price_income_table,
  aes(
    x = Year,
    y = Canterbury_PriceIncome
  )
) +
  geom_line(
    colour = "#D55E00",
    linewidth = 1.1
  ) +
  geom_point(
    colour = "#D55E00",
    size = 2.5
  ) +
  labs(
    title =
      "Canterbury House-Price-to-Income Proxy",
    subtitle =
      "Median sale price divided by mean Canterbury income",
    x = "Year",
    y =
      "Price-to-income ratio"
  ) +
  theme_minimal(
    base_size = 13
  )
```

### Interpretation

The Canterbury price-to-income proxy increased from approximately
**7.90 in 2020** to approximately **10.33 in 2022**.

It subsequently declined to approximately **9.00 by 2025**.

This indicates that purchase affordability deteriorated sharply during
the 2020–2022 housing boom before partially improving.

---

## 5.7 House-price growth vs rent growth

```{r analysis3-rent-price-plot}
price_rent_growth <- analysis3_monthly %>%
  select(
    Date,

    `House-price growth` =
      Canterbury_Price_YoY,

    `Rent growth` =
      Canterbury_Rent_YoY
  ) %>%
  pivot_longer(
    -Date,
    names_to = "Measure",
    values_to = "Growth"
  )

ggplot(
  price_rent_growth,
  aes(
    x = Date,
    y = Growth,
    colour = Measure
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    colour = "grey50"
  ) +
  geom_line(
    linewidth = 1
  ) +
  labs(
    title =
      "Canterbury House-Price Growth vs Rent Growth",
    subtitle =
      "Year-on-year percentage change",
    x = NULL,
    y = "Annual growth (%)",
    colour = NULL
  ) +
  theme_minimal(
    base_size = 13
  ) +
  theme(
    legend.position = "bottom"
  )
```

---

## 5.8 Housing-market periods

```{r analysis3-periods}
period_summary <- analysis3_monthly %>%
  group_by(
    Period
  ) %>%
  summarise(
    Mean_Price_Growth =
      mean(
        Canterbury_Price_YoY,
        na.rm = TRUE
      ),

    Mean_Rent_Growth =
      mean(
        Canterbury_Rent_YoY,
        na.rm = TRUE
      ),

    .groups = "drop"
  )

kable(
  period_summary,
  digits = 2,
  col.names = c(
    "Period",
    "House-Price Growth (%)",
    "Rent Growth (%)"
  ),
  caption =
    "Canterbury Housing Growth by Period"
)
```

### Interpretation

Average Canterbury house-price growth was approximately:

- **1.95% during 2015–2019**
- **14.82% during 2020–2022**
- **1.47% from 2023 onward**

Average rent growth was approximately:

- **-0.16% during 2015–2019**
- **6.60% during 2020–2022**
- **4.83% from 2023 onward**

The nature of affordability pressure therefore changed over time.

During 2020–2022, rapidly rising purchase prices were the dominant
problem.

After the housing boom, house-price growth slowed substantially while
rental growth remained elevated.

---

## 5.9 Rent-growth models

To keep the final analysis concise, only the immediate relationship and
the strongest individual lag relationship are retained.

```{r analysis3-models}
rent_current_model <- lm(
  Canterbury_Rent_YoY ~
    Canterbury_Price_YoY,
  data = analysis3_monthly
)

rent_lag12_model <- lm(
  Canterbury_Rent_YoY ~
    PriceGrowth_Lag12,
  data = analysis3_monthly
)
```

---

## 5.10 Newey-West rent results

```{r analysis3-newey-west}
analysis3_robust <- bind_rows(

  nw_tidy(
    rent_current_model,
    "Current relationship"
  ),

  nw_tidy(
    rent_lag12_model,
    "12-month lag"
  )
) %>%
  mutate(
    term = dplyr::recode(
      term,
      "(Intercept)" =
        "Intercept",

      "Canterbury_Price_YoY" =
        "Current house-price growth",

      "PriceGrowth_Lag12" =
        "House-price growth (12m lag)"
    )
  )

kable(
  analysis3_robust,
  digits = 4,
  col.names = c(
    "Model",
    "Variable",
    "Estimate",
    "Robust SE",
    "t-statistic",
    "p-value"
  ),
  caption =
    "Newey-West Results for Canterbury Rent Growth"
)
```

### Interpretation

The same-period relationship between house-price growth and rent growth
was positive but only weakly significant after correcting for serial
correlation.

The 12-month lag produced a stronger result.

The estimated 12-month coefficient was approximately:

**0.217**

with:

**p = 0.0005**.

This suggests that a **10 percentage-point increase in annual
house-price growth was associated with approximately 2.17 percentage
points higher annual rent growth one year later**.

This relationship should not be interpreted as proof that rising house
prices directly cause higher rents.

---

# 6. Overall Findings

The analysis produces three main findings.

First, Canterbury has historically been more affordable than New Zealand
overall.

Both Canterbury's house value-to-income multiple and mortgage-payment
burden were significantly lower than the corresponding New Zealand
measures.

Second, Canterbury's housing market remains strongly connected to the
national housing cycle.

National house-price growth was the strongest and most consistent
predictor in the final Canterbury house-price model.

Mortgage rates and credit conditions displayed meaningful standalone
relationships with Canterbury price growth, but these relationships
weakened once broader national housing conditions were controlled for.

Third, affordability deteriorated sharply during the 2020–2022 housing
boom.

House prices increased substantially faster than household income.

Although purchase-price growth later slowed, rental growth remained
relatively elevated.

The regression evidence also suggests that rent growth may respond to
earlier housing-market conditions with a delay.

---

# 7. Limitations

The analysis has several limitations.

The data are observational, so the statistical relationships cannot be
interpreted as causal effects.

Some explanatory variables, including migration and monetary-policy
variables, are measured nationally rather than specifically for
Canterbury.

Time-series models displayed serial correlation.

Newey-West robust standard errors were therefore used for final
statistical inference.

Moderate multicollinearity was present between some explanatory
variables, although VIF values remained below approximately five.

The Bright-Line variable represents different policy regimes over time
and should not be interpreted as an isolated causal estimate of the
policy's effect.

Dwelling consents represent approvals rather than completed housing
supply.

Finally, Canterbury income is available at a lower frequency than the
monthly housing-market measures.

Income was therefore analysed annually rather than artificially filling
missing monthly observations.

---

# 8. Conclusion

Canterbury has historically maintained a housing-affordability advantage
relative to New Zealand overall.

However, Canterbury still experienced substantial affordability
deterioration during the early-2020s housing boom.

House prices increased much faster than household income during
2020–2022.

Although house-price growth subsequently slowed, rental costs continued
to rise relatively strongly.

Canterbury house-price movements are closely connected to the wider New
Zealand housing market, while borrowing conditions, credit availability,
migration, housing supply and market activity provide additional
context.

Overall, Canterbury housing affordability appears to reflect a
combination of national housing-market conditions, household income,
financing costs, housing supply and rental-market pressures rather than
a single explanatory factor.
```